---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(rcea)
```

# Halpern CEA workflow

This vignette focuses on the Halpern et al. (2008) cumulative effects assessment (CEA) method using the updated `rcea` workflow. The examples use the built‑in example rasters in `inst/extdata`.

## Data preparation

You can use the catalog + cube helpers to keep workflows consistent and reproducible.

```{r}
layers_path <- system.file("extdata/catalog/layers.csv", package = "rcea")
groups_path <- system.file("extdata/catalog/groups.yaml", package = "rcea")

catalog <- load_catalog(layers_path, groups_path)

# Sensitivity matrix must match VC (rows) and driver (cols) names
sens <- matrix(
  c(
    0.8, 0.5,
    0.2, 0.7
  ),
  nrow = 2,
  dimnames = list(
    c("cod", "salmon"),
    c("shipping", "climate")
  )
)

cube <- make_cube(catalog, sensitivity = sens)
cube <- stack_layers(cube)

cube
```

At this point you have two stacks:

```{r}
names(cube$stack$drivers)
names(cube$stack$vc)
```

## CEA (Halpern) analysis

Compute cumulative effects with the fast matrix engine (default):

```{r}
ce <- cea(cube$stack$drivers, cube$stack$vc, cube$sensitivity)
ce
```

## Exposure (overlap)

```{r}
expo <- exposure(cube$stack$drivers, cube$stack$vc)
expo
```

## Post‑processing: extract and aggregate layers

The post‑processing helpers work on both `SpatRaster` outputs and matrix outputs.

### Extract (subset) layers

```{r}
# Keep only one driver (use catalog-driven names)
driver_keep <- names(cube$stack$drivers)[1]
shipping_only <- layers_extract(ce, drivers = driver_keep)
shipping_only
```

### Aggregate layers

```{r}
# Aggregate all selected layers into one
cea_total <- layers_aggregate(ce, by = "both")
cea_total
terra::plot(cea_total)

# Aggregate over drivers (one layer per VC)
cea_by_vc <- layers_aggregate(ce, by = "drivers")
cea_by_vc
terra::plot(cea_by_vc)
```

You can also change the aggregation function (e.g., mean):

```{r}
cea_by_vc_mean <- layers_aggregate(ce, by = "drivers", fun = "mean")
cea_by_vc_mean
terra::plot(cea_by_vc_mean)
```

## Per‑area effects

`layers_per_area()` computes per‑area effects using true cell sizes and returns `units` columns.

```{r}
cea_per_area <- layers_per_area(ce, cube$stack$vc)
cea_per_area
```

## Matrix output (optional)

If you want fast aggregation without repeated raster I/O, you can work in matrix space and reconstruct rasters only when needed.

```{r}
ce_mat <- cea(cube$stack$drivers, cube$stack$vc, cube$sensitivity, exportAs = "matrix")

# Aggregate in matrix space
cea_by_vc_mat <- layers_aggregate(ce_mat, by = "drivers", exportAs = "matrix")
head(cea_by_vc_mat)
```
