---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(rcea)
```

# Halpern CEA workflow

This vignette focuses on the Halpern et al. (2008) cumulative effects assessment (CEA) method using the updated `rcea` workflow. The examples use the built‑in example rasters in `inst/extdata`.

## Data preparation

You can use the catalog + cube helpers to keep workflows consistent and reproducible.

```{r}
layers_path <- system.file("extdata/catalog/layers.csv", package = "rcea")
groups_path <- system.file("extdata/catalog/groups.yaml", package = "rcea")

catalog <- load_catalog(layers_path, groups_path)

# Use sensitivity from catalog when available; otherwise define manually.
sens <- catalog$sensitivity
if (is.null(sens)) {
  sens <- matrix(
    c(
      0.8, 0.5,
      0.2, 0.7
    ),
    nrow = 2,
    dimnames = list(
      c("cod", "salmon"),
      c("shipping", "climate")
    )
  )
}

cube <- make_cube(catalog, sensitivity = sens)
cube <- stack_layers(cube)

cube
```

At this point you have two stacks:

```{r}
names(cube$stack$drivers)
names(cube$stack$vc)
```

## CEA (Halpern) analysis

Compute cumulative effects with the fast matrix engine (default):

```{r}
ce <- cea(cube$stack$drivers, cube$stack$vc, cube$sensitivity)
ce
```

### Optional preprocessing before CEA

You can preprocess layers after stacking and before CEA. This supports global
defaults and optional per-layer catalog overrides (`transform`, `normalize`).

```{r, eval=FALSE}
cube <- preprocess_layers(
  cube,
  drivers_transform = "log1p",
  drivers_normalize = "minmax",
  vc_transform = "none",
  vc_normalize = "none",
  use_catalog_overrides = TRUE
)

ce <- cea(cube$stack$drivers, cube$stack$vc, cube$sensitivity)
```

## Exposure (overlap)

```{r}
expo <- exposure(cube$stack$drivers, cube$stack$vc)
expo
```

## Post‑processing: extract and aggregate layers

The post‑processing helpers work on both `SpatRaster` outputs and matrix outputs.

### Extract (subset) layers

```{r}
# Keep only one driver (use catalog-driven names)
driver_keep <- names(cube$stack$drivers)[1]
shipping_only <- layers_extract(ce, drivers = driver_keep)
shipping_only
```

### Aggregate layers

```{r}
# Aggregate all selected layers into one
cea_total <- layers_aggregate(ce, by = "both")
cea_total
terra::plot(cea_total)

# Aggregate over drivers (one layer per VC)
cea_by_vc <- layers_aggregate(ce, by = "drivers")
cea_by_vc
terra::plot(cea_by_vc)
```

You can also change the aggregation function (e.g., mean):

```{r}
cea_by_vc_mean <- layers_aggregate(ce, by = "drivers", fun = "mean")
cea_by_vc_mean
terra::plot(cea_by_vc_mean)
```

### Group-balanced aggregation

When one pressure group has many more member layers than another (e.g., many
shipping categories vs fewer fishing categories), direct summation can over-weight
the larger group. `layers_aggregate_groups()` uses a two-stage strategy:
within-group aggregation first, then across-group aggregation.

```{r, eval=FALSE}
cea_balanced <- layers_aggregate_groups(
  ce,
  group_col = "driver_group",
  time_col = "driver_month",
  within_fun = "mean",
  across_fun = "sum"
)

# Optional group weights in across-group stage
cea_balanced_w <- layers_aggregate_groups(
  ce,
  group_col = "driver_group",
  time_col = "driver_month",
  within_fun = "mean",
  across_fun = "sum",
  group_weights = c(shipping = 0.8, fishing = 1.0)
)
```

## Per‑area effects

`layers_per_area()` computes per‑area effects using true cell sizes and returns `units` columns.

```{r}
cea_per_area <- layers_per_area(ce, cube$stack$vc)
cea_per_area
```

## Matrix output (optional)

If you want fast aggregation without repeated raster I/O, you can work in matrix space and reconstruct rasters only when needed.

```{r}
ce_mat <- cea(cube$stack$drivers, cube$stack$vc, cube$sensitivity, exportAs = "matrix")

# Aggregate in matrix space
cea_by_vc_mat <- layers_aggregate(ce_mat, by = "drivers", exportAs = "matrix")
head(cea_by_vc_mat)
```

## Temporal pairing with mixed temporal/atemporal layers

`pair_by` lets you avoid invalid cross-time combinations. By default, missing keys are
treated as broadcast (`pair_missing = "broadcast"`), so atemporal layers still apply to
all temporal slices.

```{r, eval=FALSE}
tmpl <- terra::rast(ncols = 2, nrows = 2, xmin = 0, xmax = 2, ymin = 0, ymax = 2, crs = "EPSG:3857")

# Drivers: one monthly layer + one atemporal layer
d_month <- tmpl
d_static <- tmpl
terra::values(d_month) <- 1
terra::values(d_static) <- 2
drivers <- c(d_month, d_static)
names(drivers) <- c("shipping_jan", "coastal_dev")

# VC: two monthly slices
v_jan <- tmpl
v_feb <- tmpl
terra::values(v_jan) <- 3
terra::values(v_feb) <- 4
vc <- c(v_jan, v_feb)
names(vc) <- c("cod_jan", "cod_feb")

driver_meta <- data.frame(
  layer = names(drivers),
  driver_id = c("shipping", "coastal"),
  month = c("2020-01", NA),
  stringsAsFactors = FALSE
)
vc_meta <- data.frame(
  layer = names(vc),
  vc_id = "cod",
  month = c("2020-01", "2020-02"),
  stringsAsFactors = FALSE
)

sens <- matrix(1, nrow = 1, ncol = 2, dimnames = list("cod", c("shipping", "coastal")))

# Default broadcast behavior:
# - cod_jan pairs with shipping_jan and coastal_dev
# - cod_feb pairs with coastal_dev (shipping_jan does not match)
ce_pair <- cea(
  drivers, vc, sens,
  exportAs = "matrix",
  driver_meta = driver_meta,
  vc_meta = vc_meta,
  pair_by = "month"
)
colnames(ce_pair$data)

# Strict mode (optional): requires both sides to have matching month
ce_pair_strict <- cea(
  drivers, vc, sens,
  exportAs = "matrix",
  driver_meta = driver_meta,
  vc_meta = vc_meta,
  pair_by = "month",
  pair_missing = "strict"
)
colnames(ce_pair_strict$data)
```
